---
title: > 
  MSc ENR ISM: \n
  Biodiversity analysis (alpha, beta, gamma)
author: "Marius King"
date: today
date-format: long
abstract: > 
  Analysis of the alpha, beta and gamma diversity of wetlands in the Gantrisch region (Switzerland) using R v4.5.0 [@RCoreTeam2025] and vegan [@OksanenEtAl2025].
editor: source
title-block-banner: true
format: 
  html:           
    toc: true 
    toc-depth: 2
    embed-resources: true
    theme: journal
execute:
  warning: false
  message: false
format-links: false
number-sections: true
shift-heading-level-by: 1
bibliography: bibliography.bib
output-file: index
---

## *setup* {.unnumbered}

```{r}
pacman::p_load("tidyverse", "vegan", "patchwork", "ggpubr")

set.seed(123)

theme_set(
  theme_minimal()+
  theme(
    strip.background = element_rect(fill = "black"),   
    strip.text = element_text(color = "white"),
    panel.spacing = unit(0.5, "lines"),
    axis.text = element_text(color = gray(0.3)),
    panel.border = element_rect(color = "black",fill=NA)
  )
)

colors <- c(
  "grazing"="#6B8E23",
  "mowing"="#A8D08D"
)
```

# Structure of community ecology data

**(a)** Import the community ecology data (vegetation data) from the file “Wet_Grasland.xlsx”. Tip: Remember that community ecology data consists of species data and header/environmental data. Consider to set the first column (Taxa names resp. Plot names) as rownames

**(b)** Check the data. How is the data structured? Are there NAs?

**(c)** Often in species data, as a convention and for better readability, cells are left empty if a species is not present at a sampling unit (here plot). But for analysis there should be a 0 if a species is not in the plot ► change the NAs to 0 in the Species data.

```{r}
wet_grasland <- read_delim("data/wet_grasland.csv") 

wet_grasland |> str()
wet_grasland |> summary()
  
wet_grasland <- wet_grasland |> 
  mutate(
    Species = factor(Species),
    across(everything(), ~replace_na(., 0))
  )
```


```{r}
wet_grasland_header <- read_delim("data/wet_grasland_head.csv") 

wet_grasland_header |> str()
wet_grasland_header |> summary() 

wet_grasland_header <- wet_grasland_header|> 
  mutate(
    Plot = factor(Plot),
    Mire_Pair = factor(Mire_Pair),
    Management = factor(Management),
    Municipality = factor(Municipality),
    Date = as.POSIXct(Date, format="%d/%m/%Y")
  )
```

# Community diversity

**(a)** For most operations/calculations on the species data we need the sampling units (here plots) as rows and the species as columns ► transpose the species data data frame.

**(b)** Calculate species richness. Hint: Use a base R function or the `specnumber` function of the package `vegan`

**(c)** Use the function `diversity` of the package `vegan` to calculate the Shannon diversity and the Shannon eveness

**(d)** Add the diversity measures calculated in (b) and (c) to the Header/Environmental Data

```{r}
wet_grasland_alpha <- wet_grasland |>
  pivot_longer(-Species, names_to = "Plot", values_to = "Abundance") |> 
  group_by(Plot) |> 
  summarise(
    shannon = diversity(Abundance),
    simpson = diversity(Abundance, index="simpson"),
    species_richness = specnumber(Abundance),
    eveness = shannon/log(species_richness)
  )

wet_grasland_alpha |> str()
```

```{r}
wet_grasland_header <- left_join(
  wet_grasland_header,
  wet_grasland_alpha,
  join_by(Plot)
) 

wet_grasland_header |> group_by(Management) |> 
  summarise(
    min = min(species_richness),
    mean = mean(species_richness),
    max = max(species_richness)
  )
```

```{r}
aov(species_richness~Management, wet_grasland_header) |> summary()
aov(eveness~Management, wet_grasland_header) |> summary()
aov(shannon~Management, wet_grasland_header) |> summary()
```


```{r}
plot_specnumber <- wet_grasland_header |>  
  ggplot(aes(Management, species_richness, fill=Management)) +
  geom_boxplot()+
  geom_jitter(width=0.2)+
  stat_compare_means(
    method = "anova", 
    label = "p.format", 
    label.x = 1.25, 
    label.y = 15+30*0.95
  )+
  scale_y_continuous(
    limits=c(15,45)
  )+
  scale_fill_manual(
    values=colors
  )+
  labs(
    x="",
    y="Species Richness"
  )+
  theme(legend.position = "none")

plot_eveness <- wet_grasland_header |>  
  ggplot(aes(Management, eveness, fill=Management)) +
  geom_boxplot()+
  geom_jitter(width=0.2)+
  stat_compare_means(
    method = "anova", 
    label = "p.format", 
    label.x = 1.25, 
    label.y = 0.6+0.3*0.95
  )+
  scale_y_continuous(
    limits=c(0.6,0.9)
  )+
  scale_fill_manual(
    values=colors
  )+
  labs(
    x="",
    y="Pielou's eveness"
  )+
  theme(legend.position = "none")

plot_shannon <- wet_grasland_header |>  
  ggplot(aes(Management, shannon, fill=Management)) +
  geom_boxplot()+
  geom_jitter(width=0.2)+
  stat_compare_means(
    method = "anova", 
    label = "p.format", 
    label.x = 1.25, 
    label.y = 2+1*0.95
  )+
  scale_y_continuous(
    limits=c(2,3)
  )+
  scale_fill_manual(
    values=colors
  )+
  labs(
    x="",
    y="Shannon index"
  )

plot_alpha <- plot_specnumber +
  plot_eveness +
  plot_shannon+
  plot_annotation(tag_levels = "A")

plot_alpha

ggsave(
  "plot_alpha.jpg",
  plot=plot_alpha,
  width=24,
  height=12,
  units = "cm"
)
```

**(e)** Calculate the observed gamma diversity separatly for both managements types (grazed, mowed)

**(f)** Calculate the mean beta diversity separately for both management types (grazed, mowed)

**(g)** Calculate the (plot-wise) beta diversity separately for both managements types (grazed, mowed)

```{r}
wet_grasland_gamma <- wet_grasland |>
  pivot_longer(-Species, names_to = "Plot", values_to = "Abundance") |> 
  left_join(
    wet_grasland_header,
    join_by(Plot)
  ) |> 
  group_by(Management,Species) |> 
  summarise(
    present = sum(Abundance)>0,
    .groups = "drop") |> 
  group_by(Management) |> 
  summarise(gamma = sum(present))

wet_grasland_gamma |> str()
```

```{r}
wet_grasland_header <- left_join(
  wet_grasland_header,
  wet_grasland_gamma,
  join_by(Management)
) 
```

```{r}
wet_grasland_beta <- wet_grasland_header |> 
  select(
    Plot,Management,shannon,species_richness,eveness,gamma
  ) |> 
  group_by(Management) |> 
  mutate(
    mean_species_richness = species_richness |> mean(),
    .groups = "drop"
  ) |> 
  mutate(
    beta_mean = gamma/mean_species_richness,
    beta = gamma/species_richness
  )

wet_grasland_beta |> summary()
wet_grasland_beta |> 
  group_by(Management) |> 
  summarise(
    beta_mean = first(beta_mean)
  )
```

```{r}
aov(beta~Management, wet_grasland_beta) |> summary()
```

```{r}
plot_beta <- wet_grasland_beta |> 
  ggplot(aes(Management, beta, fill=Management)) +
  geom_boxplot()+
  geom_jitter(width=0.2)+
  stat_compare_means(
    method = "anova", 
    label = "p.format", 
    label.x = 1.25, 
    label.y = 2.1+3*0.95
  )+
  scale_y_continuous(
    limits=c(2.1,5.1)
  )+
  scale_fill_manual(
    values=colors
  )+
  labs(
    x="",
    y="Beta diversity"
  )

plot_beta +
  plot_annotation(tag_levels = "A")
```

# References {.unnumbered}

::: {#refs}
:::


